---
title: "Homework 3"
author: "wendyxwang"
date: "Tuesday, February 24, 2015"
output: html_document
reader: Ning Li
---

#Homework 3 Questions:
Reproduce the results presented in Figure 2 of the following paper: Qian, F., Bolen, C. R., Jing, C., Wang, X., Zheng, W., Zhao, H., et al. (2013). Impaired toll-like receptor 3-mediated immune responses from macrophages of patients chronically infected with hepatitis C virus. Clinical and Vaccine Immunology : CVI, 20(2), 146–155. doi:10.1128/CVI.00530-12

You will have to:

Get the data from GEO
Normalize the data (if necessary)
Use limma to test for differential expression
Display the results using a heatmap [Hint: Use the pheatmap package]


#Answers:
First, set up some general options and get the packages ready

```{r}
library("knitr")
opts_chunk$set(tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=60), cache=TRUE, messages=FALSE)

library(limma)
library(Biobase)
library(data.table)
library(statmod)

#set working directory
#setwd("E:/Course2015Winter/BIOST578/RStudio clone/HW3_Xiaoliang_Wang")
```


#1. Get data from GEO
```{r}
library(GEOquery)

#Get the data that was used in the paper
GSE40812_set <- getGEO("GSE40812", destdir = "Data/GEO/")[[1]]
#somehow my code was using the soft file, so I need to force it to use the matrix.gz file

#check out the eSet
str(GSE40812_set,max.level=2)
class(GSE40812_set)
exprs(GSE40812_set)[1:3,1:4]

pData(GSE40812_set)[1:2, ]
```


Clean the pData
```{r}
gset <- GSE40812_set
#Sanitize data
sanitize_pdata <- function(pd){
keepCols <- c("title", "geo_accession", "source_name_ch1", "characteristics_ch1", "characteristics_ch1.2")
pd <- pd[, keepCols]
colnames(pd) <- c("ptid","GSM_id","source", "infection_status","treatment")
#we need to substract the id from title for paired comparison later
pd$ptid <- sapply(strsplit(as.character(pd$ptid),"_"), function(x) x[[2]])
pd$infection_status <- gsub(".*: ", "", pd$infection_status)
pd$treatment <- gsub(".*: ", "", pd$treatment)

pd
}

pData(gset) <- sanitize_pdata(pData(gset))

pData(gset)[1:10,]
```

We also want to make sure we are only using the Macrophage data
```{r}
gset_m <- gset
macrophage <- pData(gset_m)[pData(gset_m)$source=="Monocyte-derived Macrophage",]
eset <- gset_m[,rownames(macrophage)]
```
Some of the data has been excluded.


#2. Normalize the data if necessary
```{r}
## From the viewer, it seems like the data has already been normalized.
#make a boxplot
boxplot(exprs(eset))
#seems really comparable
## So, we do not need to normalize the data
```

#3. Use limma to test for differential expression
```{r}
#design matrix
mm <- model.matrix(~treatment, data=eset)
#lmfit
fit <- lmFit(eset,mm)
ebay <- eBayes(fit)
```

Test for the treatment effect
```{r}
colnames(fit$coef)

#We have mock as reference treatment, and poly I:C(H) as the other treatment group
```

**Poly I:C(HMW) vs. MOCK**
```{r}
topPoly_H <- topTable(ebay, coef="treatmentPoly IC H", number=Inf, sort.by="none")
colnames(topPoly_H)

#Accoridng to the paper, use the cutoff values of 1.5 fold change and 0.05 FDR
summ_topPoly_H <- sum(topPoly_H$adj.P.Val<0.05 & abs(topPoly_H$logFC)>=log2(1.5))
```
This returns 1146 significant results at the threshold that we selected.

```{r}
summ_topPoly_H_pos <- sum(topPoly_H$adj.P.Val<0.05 & topPoly_H$logFC>=log2(1.5))
```
753 upregulated responses, and 393 downregulated reponses, which is the same as reported in the original paper.


**We then limit our analysis on the responsive genes**
```{r}
eset_r <- eset[topPoly_H$adj.P.Val<0.05 & abs(topPoly_H$logFC)>log2(1.5),]
pData(eset_r)$treatment <- ifelse(eset_r$treatment=="Mock", "Mock", "Poly")
infect_treat <- factor(paste(eset_r$infection_status,eset_r$treatment,sep=""))

#re-define design matrix
mm_r<-model.matrix(~0+infect_treat, data=eset_r)
colnames(mm_r)<-levels(infect_treat)

corfit<-duplicateCorrelation(eset_r, mm_r, block=pData(eset_r)$ptid)
corfit$consensus

fit_r<-lmFit(eset_r,mm_r, block=pData(eset_r)$ptid, correlation=corfit$consensus)
ebay_r<-eBayes(fit_r)
colnames(fit_r)
```

We also need to definte contrast manually
```{r}
#specify contrast
cm <- makeContrasts(
  PolyMockforPos=PosPoly-PosMock, 
  PolyMockforNeg=NegPoly-NegMock,
  PolyforPosNeg=PosPoly-NegPoly,
  MockforPosNeg=PosMock-NegMock,
  All=(PosPoly-PosMock)-(NegPoly-NegMock),
  levels=mm_r)

#Re-fit the model with specified contrast
fit_r2<- contrasts.fit(fit_r,cm)
ebay_r2<- eBayes(fit_r2)
```


**Check the significant genes**
```{r}
top_r <- topTable(ebay_r2, coef="All", number=Inf, sort.by="none")

summ_top_adj <- sum(top_r$adj.P.Val<0.1)
#no genes were found to be significantly different following FDR corrections

summ_top_r <- sum(top_r$P.Val<0.1)
#we have 41 probe sets returned.
```
Not sure why we have 2 probe sets fewer than reported in the paper


#4. Display the results using a heatmap
```{r}
library(pheatmap)

#Subset the probes with significant difference in the second analysis
sign <- eset_r[top_r$P.Val<0.1,]
```

Draw heatmap
```{r}
sign_pd <- pData(sign)
sign_pd<- sign_pd[order(sign_pd$treatment, sign_pd$infection_status, sign_pd$ptid), ]
hm_colorder <- rownames(sign_pd)
hm_colnames <- sign_pd$ptid
hm_data <- exprs(sign)
hm_dt <- as.data.table(hm_data)
rownames(hm_dt) <- rownames(hm_data)
setcolorder(hm_dt, hm_colorder)
hm_data <- data.matrix(hm_dt)
colnames(hm_data) <- hm_colnames
heatmap(hm_data, Colv=NA)
```


##Citation
1. Dr. Gottardo BIOST 578A Differential Expression slides
2. LIMMA userguide

